Modal Abstractions for Virtualizing Memory Addresses

### 2025-07-29 examples_C.tex
bus draft 25 pages

\subsubsection{From L$n+1$ Entries to L$n$ Tables}
We will discuss access to the level 4 table later (Section \ref{wlkpgdirC}). However, for subsequent levels, the base address of level $n$ must be
fetched from the appropriate entry in the table of level $n+1$.
This is the role of \lstinline|pte_get_next_table| (originally Figure \ref{fig:pagetablescode}, with proof details in Figure \ref{fig:calltopteinitializeC}).
It is passed the virtual address of the page table entry in level $n+1$, and should return the \emph{virtual} 
address of the \emph{base} of the level $n$ table
indicated by that entry.
If the entry is empty (i.e., this is a sparse part of the page table representation),
the code also allocates a page for the level $n$ table, installs it in the level $n+1$ entry, and establishes appropriate invariants.
Figure \ref{fig:calltopteinitializeC} presents the function with proof annotations that we will explain shortly, but we first explain the functionality.
\lstinline|pte_get_next_table| accepts a \emph{virtual} address \lstinline|entry| which points to a level-$n+1$
table entry.

On Line \ref{line:read_entry_contentsC}, the code checks the present bit of the entry.
If the bit is unset, there is no level-$n$ table, so one must be allocated via \lstinline|pte_initialize| (explained shortly\footnote{
  Briefly, it allocates a fresh physical page, and initializes the memory pointed to by \lstinline|entry| with that physical address.
}) and marked present.
By Line \ref{line:finalpieceS} the entry is known to be valid and contain the physical address of
the base of a level $n$ table. That address is then extracted (Line \ref{line:extract_pfn}),
converted to a virtual address (Line \ref{line:p2vC}), and returned to the caller.
We can now discuss \lstinline|pte_get_next_table|'s proof of correctness.
While at first glance this code does not sound terribly complex, beyond needing some care to distinguish physical and virtual addresses,
it has a highly nontrivial correctness argument, which depends critically on detailed invariants on how access to page table
entries is shared between parts of the kernel. No prior work has engaged with this problem.

@@

Even given the slight adaptation of our assembly-level proof for the C-level presentation in Figure \ref{walkpgdirC},
the proof outline in the figure omits may assertions that hold at various points in time.
But by Line \ref{line:ex_l3_vpteC}, it should be clear that the proof is on track to accumulate a set of similar assertions for each level.
Figure \ref{fig:rwalkC} expands the abbreviated postcondition to include the full set of facts that are accumulated in this way.
$\mathsf{R}_\mathsf{walk}$ together with $\mathsf{R}_{\mathsf{l1e}}$ should look quite similar to \lstinline|L4_L1_PointsTo| (Figure \ref{fig:strongvirtualpointsto}),
and this is quite intentional. \lstinline|walkpgdir|'s execution observes enough
to determine that there is most of an address translation for virtual address \lstinline|va|,
at least down to a possibly-invalid L1 entry (which \lstinline|walkpgdir|'s caller, \lstinline|vaspace_mappage|, will check). 
Each virtual pte-points-to in $\mathsf{R}_\mathsf{walk}$ internally contains the physical points-to portion of one page table walk step for \lstinline|L4_L1_PointsTo|,
and the pure assertions in $\mathsf{R}_\mathsf{walk}$ ensure the address arithmetic works.
$\mathsf{R}_{\mathsf{l1e}}$ also includes the self-conditional fractional ownership of the L1 entry (per Section \ref{sec:selfconditional})
for the caller to initialize the entry if it is empty, making it possible for the caller to complete virtual points-to assertions
for a newly-mapped page, as we discuss next.

@@

For brevity this example is specialized to the case where $\vaddr$ is known to not be mapped:
the precondition on Line \ref{line:mappage_pre} includes $\theta \; !!\; \vaddr = \texttt{None}$.\footnote{
  The generalization to handling either case and returning an error is straightforward.
}
The precondition on Line \ref{line:mappage_pre}, directly entails
the precondition of the \lstinline|walkpgdir| call.\footnote{The proof of \lstinline|vaspace_mappage|'s caller
would extract this single identity map token for the specific L4 entry from a set of 512 that are
part of the kernel invariant, as \lstinline|walkpgdir|'s proof does for lower levels.
}
\lstinline|walkpgdir|, as just discussed,
returns the virtual address of an allocated L1 entry and its postcondition contains almost all of the information
needed to construct a virtual points-to for \lstinline|va| --- except information about the L1 entry
being present and pointing to a data page.
We already discussed for the upper level page-tables how the entry-present checks are handled, and
Line \ref{line:mappage_pte_present_startC} is similar: $\mathsf{R}_{\textsf{l1e}}$ includes the self-conditional
fractional permission for the L1 entry, so as it is not present, by Line \ref{line:l1entry_storeC}
it is known that full permission is held to update that entry.
Lines \ref{line:l1entry_storeC} and \ref{line:l1entry_setpresent} are verified using the pte-points-to
memory store rule.\footnote{In the assembly, \lstinline|fpaddr| is already known to be word-aligned, so it is simply bitwise-and-ed with 1
(setting the present flag) and a single store occurs.}
\looseness=-1

>>>>

\subsubsection{From L$n+1$ Entries to L$n$ Tables}
We discuss access to the level 4 table later (Section \ref{wlkpgdirC}). However, for subsequent levels, the base address of level $n$ must be
fetched from the appropriate entry in the table of level $n+1$.
This is the role of \lstinline|pte_get_next_table| (originally Figure \ref{fig:pagetablescode}, with proof details in Figure \ref{fig:calltopteinitializeC}).
It is passed the virtual address of the page table entry in level $n+1$, and should return the \emph{virtual} 
address of the \emph{base} of the level $n$ table
indicated by that entry.
If the entry is empty (i.e., this is a sparse part of the page table representation),
the code also allocates a page for the level $n$ table, installs it in the level $n+1$ entry, and establishes appropriate invariants.
Figure \ref{fig:calltopteinitializeC} presents the function with proof annotations that we will explain shortly, but we first explain the functionality.
\lstinline|pte_get_next_table| accepts a \emph{virtual} address \lstinline|entry| which points to a level-$n+1$
table entry.
\looseness=-1

On Line \ref{line:read_entry_contentsC}, the code checks the present bit of the entry.
If the bit is unset, there is no level-$n$ table, so one must be allocated via \lstinline|pte_initialize| (explained shortly\footnote{
  Briefly, it allocates a fresh physical page, and initializes the memory pointed to by \lstinline|entry| with that physical address.
}) and marked present.
By Line \ref{line:finalpieceS} the entry is known to be valid and contain the physical address of
the base of a level $n$ table. That address is then extracted (Line \ref{line:extract_pfn}),
converted to a virtual address (Line \ref{line:p2vC}), and returned to the caller.
We can now discuss \lstinline|pte_get_next_table|'s proof of correctness.
While at first glance this code may look like its subtlety is mostly care to distinguish physical and virtual addresses,
it has a highly nontrivial correctness argument, which depends critically on detailed invariants on how access to page table
entries is shared between parts of the kernel. No prior work has engaged with this problem.
\looseness=-1

@@

Even given the slight adaptation of our assembly-level proof for the C-level presentation in Figure \ref{walkpgdirC},
the proof outline in the figure omits some repeat intermediate assertions for readability.
But by Line \ref{line:ex_l3_vpteC}, it should be clear that the proof accumulates a set of similar assertions for each level.
Figure \ref{fig:rwalkC} expands the abbreviated postcondition to the full set of facts that are accumulated in this way.
$\mathsf{R}_\mathsf{walk}$ together with $\mathsf{R}_{\mathsf{l1e}}$
nearly entail \lstinline|L4_L1_PointsTo| (Figure \ref{fig:strongvirtualpointsto}) within the logic,
forming the basis of the construction of a new virtual points-to for virtual address \textsf{va}.
\lstinline|walkpgdir|'s execution observes most evidence of an address translation for \lstinline|va|,
at least down to a possibly-invalid L1 entry (which \lstinline|walkpgdir|'s caller, \lstinline|vaspace_mappage|, will check). 
Each virtual pte-points-to in $\mathsf{R}_\mathsf{walk}$ internally contains the physical points-to portion of one page table walk step for \lstinline|L4_L1_PointsTo|,
and the pure assertions in $\mathsf{R}_\mathsf{walk}$ ensure the address arithmetic works.
$\mathsf{R}_{\mathsf{l1e}}$ includes the self-conditional fractional ownership of the L1 entry (Section \ref{sec:selfconditional})
for the caller to initialize the entry if it is empty, so the caller can complete a virtual points-to assertions
for a newly-mapped page, as we discuss next.
\looseness=-1

@@

For brevity this example is specialized to the case where $\vaddr$ is known to not be mapped:
the precondition on Line \ref{line:mappage_pre} includes $\theta \; !!\; \vaddr = \texttt{None}$.\footnote{
  The generalization to handling either case and returning an error is straightforward.
}
The precondition on Line \ref{line:mappage_pre}, directly entails
the precondition of the \lstinline|walkpgdir| call.\footnote{The proof of \lstinline|vaspace_mappage|'s caller
would extract this single identity map token for the specific L4 entry from a set of 512 that are
part of the kernel invariant, as \lstinline|walkpgdir|'s proof does for lower levels.
}
\lstinline|walkpgdir|, as just discussed,
returns the virtual address of an allocated L1 entry and its postcondition contains almost all of the information
needed to construct a virtual points-to for \lstinline|va| --- except information about the L1 entry
being present and pointing to a data page.
We already discussed for the upper level page-tables how the entry-present checks are handled, and
Line \ref{line:mappage_pte_present_startC} is similar: $\mathsf{R}_{\textsf{l1e}}$ includes the self-conditional
fractional permission for the L1 entry, so as it is not present, by Line \ref{line:l1entry_storeC}
it is known that full permission is held to update that entry.
Lines \ref{line:l1entry_storeC} and \ref{line:l1entry_setpresent} are verified using the pte-points-to
memory store rule.\footnote{In the assembly proof, this is a bitwise-and of the word-aligned\lstinline|fpaddr| with 1
(setting the present flag), yielding a single.
\looseness=-1
}
\looseness=-1

### 2025-07-29 relwork.tex
bus draft 25 pages

\subsection{OS Kernel Verification}
There has been relatively little prior work on formal verification of virtual memory.
\replace{Instead, m}Most OS verification work \replace{has focused on minimizing}{minimizes} reasoning about virtual memory management.
The original \textsc{Verisoft} project~\cite{alkassar2008verisoft,alkassar2010pervasive,alkassar2008formal,dalinger2005verification,hillebrand2005address,alkassar2008formal,starostin2010formal} 
relied on custom hardware\del{,} which\del{, among other things,} always ran kernel code with virtual memory disabled, removing the circularity that is a key challenge of verifying
VMM code for real hardware: \replace{A}{a}t that point page tables become a basic partial map data structure to represent user program address translations,
with an idiosyncratic format. \replace{It turns out that s}{S}ubsequent OS verification work
\add{targets real hardware that uses virtual addressing in the kernel, but unsoundly
uses hardware models that do not.} \del{also treats page tables this way, but
unsoundly given that \replace{other projects}{their} target hardware \del{that} \emph{does} run the kernel with address translation:}
\add{Thus} they \emph{trust}
that the particular page table manipulations do not, for example, unmap kernel code\footnote{\add{A ``fun'' bug one author encountered in
an industrial prototype kernel.}}
 (which can crash the machine even if done ``temporarily'').
% \footnote{A bug
% one author has personally encountered in a research kernel, though not one discussed here.}).
This is true for \textsc{seL4}~\cite{Klein2009seL4,seL4TOCS,Sewell2013translation}, whose formal machine model omits address translation,
and \textsc{CertiKOS}~\cite{gu15,gu2016certikos,gu2018certikos,chen2016interrupts}, whose refinement proofs rely on
\textsc{CompCert}'s usermode-oriented memory model~\cite{leroy2008formal,leroy2009formally} which assumes
updates to one memory address are independent of updates to another ---
which is not true of page table updates.
Other work on OS verification either never progressed \del{far enough} to \del{address} VMM verification
(\textsc{Verisoft XT}~\cite{cohen2009vcc,cohen2010local,dahlweid2009vcc,cohen2013SOFSEM}),
or uses memory-safe languages
\replace{to enable safe cohabitation of a single address space by all processes}{for process isolation instead of address translation}
(\textsc{Singularity}~\cite{Fahndrich2006language,Hunt2007singularity,Hunt2007sealing,Barnett2011specsharp}, \textsc{Verve}~\cite{Yang2010Verve},
and \textsc{Tock}~\cite{levy2017multiprogramming})\replace{. The latter approach ensures}{, ensuring} memory safety, but ignor\replace{es}{ing} other functional uses of virtual memory
hardware, like swapping~\cite{Denning1970VM} or exploiting copy-on-write techniques for dynamic migration of virtual machines~\cite{clark2005live}.
\looseness=-1

\add{
 These limitations motivate work like ours on reasoning soundly about virtual memory management.
 As discussed earlier, 
}
\citet{kolanski08vstte} are the only other researchers to study VMM verification against a realistic hardware
model, where page table updates are performed through virtual memory accesses (later adding C-level type modeling~\cite{kolanski09tphols}).
As noted in Section \ref{sec:overly-restrictive}, Kolanski and Klein's virtual points-to is similar to that in Figure \ref{fig:strongvirtualpointsto},
with the attendant problems discussed earlier and lifted by our model. Their approach had modal elements,
but
\replace{the work never addressed the issues of managing multiple address spaces, switching address spaces, or
invariants referring to other address spaces, which our explicitly modal approach and case studies do address.
They verified a fraction of an ARM version of our Figure ??, which assumed that they already had the virtual address
of the L1 entry (which they assumed existed), thereby skipping the physical-to-virtual conversion issues we treat in Section ??.
}{
 did not tackle evaluations that would benefit from modality.
}
\looseness=-1

\add{
 Our work improves significantly on the technical capabilities of ther logic and
 evaluates on kernel code that is more complete and more challenging than theirs.
 Our modal approach makes it possible to specify address space changes cleanly, which their logic cannot do at all.
 Our use of virtual pte-points-to assertions enables nearly the same proof rules\todo{if this stays, we should typeset one of the vpte-pointsto rules to justify this}
 as standard memory accesses, and constructing virtual points-to information within the logic
 (c.f.\ the logical entailment between $\textsf{R}_\textsf{walk}$ and $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$)
 rather than reasoning semantically about when the model state supports new virtual points-to assertions.
}

\add{
While Kolanski and Klein verify the critical step of updating an already-located L1 entry to map a new page
(ARM assembly corresponding roughly to Lines \ref{line:l1entry_storeC} and \ref{line:l1entry_setpresent}
in our Figure \ref{fig:mapping_codeC}), they ignore the essential code
preceding that step --- which as our \lstinline|walkpgdir| and \lstinline|pte_get_next_table| verifications
demonstrate, side-steps a significant amount of complexity and critical reasoning tasks.
We have verified the entirety of the software page table walk up to mapping a new page, aside from a trusted physical memory allocator.\footnote{
 \add{Which works quite similarly to verified \lstinline|malloc|~\cite{Chlipala2013Bedrock,wickerson2010explicit}}
}
 As a consequence of tackling this larger verification challenge, we are also the first to formally specify large
 segments of the self-referential portion of an OS kernel's virtual memory management invariants (per Sections \ref{sec:p2vC} and \ref{sec:selfconditional}),
 and to reason about converting back from physical addresses to virtual addresses efficiently.
}

@@

\add{
 No prior work has dealt with combined address space changes and context switching. \textsc{XCAP}~\cite{ni2007contexts} deals
 only with usermode threading (in a single process). \textsc{CertiKOS} and \textsc{seL4} trust assembly primitives for
 context switches, in addition to not modeling address translation for executing code.
}
\todo[inline]{author response talks about syeda and klein here briefly, and it fits, but we also want the longer discussion}

\paragraph{Translation Lookaside Buffers}
As noted in Section \ref{sec:backgroundonmachinemodel}, we do not formally model or reason about TLBs.
\replace{TLB flushes are required only when addresses are removed from a virtual address space, or when changing
virtual address spaces.}{
 TLB flushes are necessary when a page is \emph{un}mapped, or when switching address spaces.
}
\replace{Because t}{T}his occurs in few places in \replace{the}{uniprocessor} kernel\add{s} (in some, only 3 locations), \add{so} full verified
kernels including \textsc{seL4}~\cite{Klein2009seL4,seL4TOCS} and \textsc{CertiKOS}~\cite{gu15,gu2016certikos}
trust TLB management. Neither of the aforementioned systems has a hardware model including a TLB, so neither is able
to verify TLB management in any form --- they \emph{must} trust its operation.
\replace{
The only place TLB code becomes particularly challenging is in multiprocessor kernels, where
not only must the running CPU flush its own TLB with special instructions, but it must send an inter-processor interrupt (IPI) to all
other cores to ensure they also flush the relevant ranges of their own TLBs.
Currently no formal hardware model exists with sufficient detail to reason about IPIs in full detail,
which would require extensions to model interactions with ACPI tables populated by firmware on desktop and server machines,
or flattened device trees for (semi-)embedded systems like single-board ARM and RISC-V machines, which tell the OS
important information about how to control other CPU cores;
plus the memory-mapped IO used to trigger those IPIs.
As a result, multicore extensions of \textsc{seL4}~[?] and \textsc{CertiKOS}~[?] also trust this functionality.
}{
 This is true of multicore verified kernels as well~\cite{von2013clustered,gu2016certikos}, though there
 the situation becomes much more challenging: when unmapping pages, the running CPU must invalidate the relevant TLB entries locally,
 but also send an inter-procesor interrupt (IPI) to all other cores to ensure they also invalidate affected entries on their TLBs.
 No formal hardware model currently exists for IPIs on any architecture, or even for the memory-mapped IO
 used to trigger those IPIs.
}

@@

\subsection{Program Logics with Modalities}
Modalities have long been a staple of program logics, at least as far back as Dijkstra's weakest precondition calculus~\cite{dijkstra1975guarded}
and Pratt's observation the Hoare triples
could be decomposed using the weakest-precondition modality of dynamic logic~\cite{pratt1976semantical},
in a form quite similar to what \iris uses today~\cite{jung2018iris}.
Variants of Nakano's later modality~\cite{nakano2000modality} have long been used to deal with step-indexing 
for impredicative and recursive features of logics and type systems~\cite{Appel2007,hobor2010theory,birkedal2011step,birkedal2013intensional,jung2018iris}.

>>>>

\paragraph{OS Kernel Verification}
There has been relatively little prior work on formal verification of virtual memory.
\replace{Instead, m}Most OS verification work \replace{has focused on minimizing}{minimizes} reasoning about virtual memory management.
The original \textsc{Verisoft} project~\cite{alkassar2008verisoft,alkassar2010pervasive,alkassar2008formal,dalinger2005verification,hillebrand2005address,alkassar2008formal,starostin2010formal} 
relied on custom hardware\del{,} which\del{, among other things,} always ran kernel code with virtual memory disabled, removing the circularity that is a key challenge of verifying
VMM code for real hardware: \replace{A}{a}t that point page tables become a basic partial map data structure to represent user program address translations,
with an idiosyncratic format. \replace{It turns out that s}{S}ubsequent OS verification work
\add{targets real hardware that uses virtual addressing in the kernel, but unsoundly
uses hardware models that do not.} \del{also treats page tables this way, but
unsoundly given that \replace{other projects}{their} target hardware \del{that} \emph{does} run the kernel with address translation:}
\add{Thus} they \emph{trust}
that the particular page table manipulations do not, for example, unmap kernel code
% \footnote{\add{A ``fun'' bug one author encountered in
% an industrial prototype kernel.}}
 (which can crash the machine even if done ``temporarily'').
% \footnote{A bug
% one author has personally encountered in a research kernel, though not one discussed here.}).
This is true for \textsc{seL4}~\cite{Klein2009seL4,seL4TOCS,Sewell2013translation}, whose formal machine model omits address translation,
and \textsc{CertiKOS}~\cite{gu15,gu2016certikos,gu2018certikos,chen2016interrupts}, whose refinement proofs rely on
\textsc{CompCert}'s usermode-oriented memory model~\cite{leroy2008formal,leroy2009formally} which assumes
updates to one memory address are independent of updates to another ---
which is not true of page table updates.
Other work on OS verification either never progressed \del{far enough} to \del{address} VMM verification
(\textsc{Verisoft XT}~\cite{cohen2009vcc,cohen2010local,dahlweid2009vcc,cohen2013SOFSEM}),
or uses memory-safe languages
\replace{to enable safe cohabitation of a single address space by all processes}{for process isolation instead of address translation}
(\textsc{Singularity}~\cite{Fahndrich2006language,Hunt2007singularity,Hunt2007sealing,Barnett2011specsharp}, \textsc{Verve}~\cite{Yang2010Verve},
and \textsc{Tock}~\cite{levy2017multiprogramming})\replace{. The latter approach ensures}{, ensuring} memory safety, but ignor\replace{es}{ing} other functional uses of virtual memory
hardware, like swapping~\cite{Denning1970VM} or exploiting copy-on-write techniques for dynamic migration of virtual machines~\cite{clark2005live}.
\looseness=-1

\add{
 These limitations motivate work like ours on reasoning soundly about virtual memory management.
 As discussed earlier, 
}
\citet{kolanski08vstte} are the only other researchers to study VMM verification against a realistic hardware
model, where page table updates are performed through virtual memory accesses (later adding C-level type modeling~\cite{kolanski09tphols}).
As noted in Section \ref{sec:overly-restrictive}, Kolanski and Klein's virtual points-to is similar to that in Figure \ref{fig:strongvirtualpointsto},
with the attendant problems discussed earlier and lifted by our model. Their approach had modal elements,
but
\replace{the work never addressed the issues of managing multiple address spaces, switching address spaces, or
invariants referring to other address spaces, which our explicitly modal approach and case studies do address.
They verified a fraction of an ARM version of our Figure ??, which assumed that they already had the virtual address
of the L1 entry (which they assumed existed), thereby skipping the physical-to-virtual conversion issues we treat in Section ??.
}{
 did not tackle evaluations that would benefit from modality.
}
\add{
 Our work improves significantly on the technical capabilities of ther logic and
 evaluates on kernel code that is more complete and more challenging than theirs.
}
\looseness=-1

\add{
 Our modal approach makes it possible to specify address space changes cleanly, which their logic cannot do at all.
 Our use of virtual pte-points-to assertions enables nearly the same proof rules
 as standard memory accesses, and constructing virtual points-to information within the logic
 (c.f.\ the logical entailment between $\textsf{R}_\textsf{walk}$ and $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$)
 whereas Kolanski and Klein must reason semantically about when the model state supports new virtual points-to assertions.
}

\add{
Kolanski and Klein verify the critical step of updating an already-located L1 entry to map a new page
(ARM assembly corresponding roughly to Lines \ref{line:l1entry_storeC} and \ref{line:l1entry_setpresent}
in our Figure \ref{fig:mapping_codeC}), but ignore the essential code
preceding that step --- which as our \lstinline|walkpgdir| and \lstinline|pte_get_next_table| verifications
demonstrate, side-steps a significant amount of complexity and critical reasoning tasks.
We have verified the entirety of the software page table walk up to mapping a new page, aside from a trusted physical memory allocator resembling
\lstinline|malloc|~\cite{Chlipala2013Bedrock,wickerson2010explicit}.
}
 As a consequence of tackling this larger verification challenge, we are also the first to formally specify large
 segments of the self-referential portion of an OS kernel's virtual memory management invariants (per Sections \ref{sec:p2vC} and \ref{sec:selfconditional}),
 and to reason about converting back from physical addresses to virtual addresses efficiently.
}
\looseness=-1

@@

\add{
 No prior work has dealt with combined address space changes and context switching. \textsc{XCAP}~\cite{ni2007contexts} deals
 only with usermode threading (in a single process). \textsc{CertiKOS} and \textsc{seL4} trust assembly primitives for
 context switches, in addition to not modeling address translation for executing code.
}
% \todo[inline]{author response talks about syeda and klein here briefly, and it fits, but we also want the longer discussion}

\del{Translation Lookaside Buffers}
As noted in Section \ref{sec:backgroundonmachinemodel}, we do not formally model or reason about \add{translation lookaside buffers} \add{(}TLBs\add{)}.
\replace{TLB flushes are required only when addresses are removed from a virtual address space, or when changing
virtual address spaces.}{
 TLB flushes are necessary when a page is \emph{un}mapped, or when switching address spaces.
}
\replace{Because t}{T}his occurs in few places in \replace{the}{uniprocessor} kernel\add{s} (in some, only 3 locations), \add{so} full verified
kernels including \textsc{seL4}~\cite{Klein2009seL4,seL4TOCS} and \textsc{CertiKOS}~\cite{gu15,gu2016certikos}
trust TLB management. Neither of the aforementioned systems has a hardware model including a TLB, so neither is able
to verify TLB management in any form --- they \emph{must} trust its operation.
\replace{
The only place TLB code becomes particularly challenging is in multiprocessor kernels, where
not only must the running CPU flush its own TLB with special instructions, but it must send an inter-processor interrupt (IPI) to all
other cores to ensure they also flush the relevant ranges of their own TLBs.
Currently no formal hardware model exists with sufficient detail to reason about IPIs in full detail,
which would require extensions to model interactions with ACPI tables populated by firmware on desktop and server machines,
or flattened device trees for (semi-)embedded systems like single-board ARM and RISC-V machines, which tell the OS
important information about how to control other CPU cores;
plus the memory-mapped IO used to trigger those IPIs.
As a result, multicore extensions of \textsc{seL4}~[?] and \textsc{CertiKOS}~[?] also trust this functionality.
}{
 This is true of multicore verified kernels as well~\cite{von2013clustered,gu2016certikos}, though there
 the situation becomes much more challenging: when unmapping pages, the running CPU must invalidate the relevant TLB entries locally,
 but also send an inter-procesor interrupt (IPI) to all other cores to ensure they also invalidate affected entries on their TLBs.
 No formal hardware model currently exists for IPIs on any architecture, or even for the memory-mapped IO
 used to trigger those IPIs.
}

@@

\paragraph{Program Logics with Modalities}
Modalities have long been a staple of program logics, at least \replace{as far back as}{since} Dijkstra's weakest precondition calculus~\cite{dijkstra1975guarded}
and Pratt's observation the Hoare triples
could be decomposed using the weakest-precondition modality of dynamic logic~\cite{pratt1976semantical},
in a form quite similar to what \iris uses today~\cite{jung2018iris}.
Variants of Nakano's later modality~\cite{nakano2000modality} have long been used to deal with step-indexing 
for impredicative and recursive features of logics and type systems~\cite{Appel2007,hobor2010theory,birkedal2011step,birkedal2013intensional,jung2018iris}.
\looseness=-1
